---
title: "Random Forest Lipid -2"
author: "Roshan Shafiha"
date: "6/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(dev = "png",
                      dpi = 300,
                      echo = TRUE,
                      cache = TRUE)
```

## Random Forest

Now the datasets would be exchanged to test and validate the model.

Train & Test set - Dataset 2  (Fenland data) 

Validation set - Dataset 1 (Italian cohort) 

load the libraries 

```{r,message=FALSE}

library(randomForest)
library(caret)
library(e1071)
library(alookr)
library(dplyr)
library(pROC)
library(ggplot2)
library(ggpubr)
library(ggsignif)
library(enrichR)
library(DT)
library(corrplot)
library(dplyr)
library(tibble)

```

Load the data

```{r }

clinical<-read.csv("clinical_data_significant.csv",header = T,row.names = 1)

data<-read.csv("lipid_data_significant.csv",header = T,row.names = 1)

data<-as.data.frame(t(data))

data<-add_column(data, sample=clinical$NAFLD, .before = 1)

head(data)

summary(data[,2:ncol(data)])

```


lets load the validation set 

```{r}
validation_lipid<-as.data.frame(read.csv("validation_lipid.csv",header = T,row.names = 1))

validation_lipid<-as.data.frame(scale(t(validation_lipid)))

validation_clinical<-read.csv("validation_clinical.csv",header = T,row.names = 1)

all(rownames(validation_lipid) == validation_clinical$ID)

validation_lipid<-add_column(validation_lipid,sample=validation_clinical$steatosis,.before = 1)

validation_lipid$sample[validation_lipid$sample == 1] <- "steatosis1"

validation_lipid$sample[validation_lipid$sample == 0] <- "steatosis0"


#extract the columns of the data present in the validation set 

data<-data[,colnames(data) %in% colnames(validation_lipid)]

head(data)

```


lets alternate the train and validate sample. Dataset 2 becomes the train set and Dataset 1 becomes the test set. 

```{r}
data_train<-validation_lipid

validation<-data

head(data_train)

head(validation)
```





lets split the dataset

```{r}
sb <- data_train %>%
  split_by(sample, seed = 6534)

attr_names <- names(attributes(sb))

sb_attr <- attributes(sb)

summary(sb)

```


extract the train and the test

```{r}
train <- sb %>%
  extract_set(set = "train")

test <- sb %>%
  extract_set(set = "test")

table(train$sample)


```


## stratified k fold cross validation 


```{r}

folds <- 5

cvIndex <- createFolds(factor(train$sample), folds, returnTrain = T)

tc <- trainControl(index = cvIndex,
               method = 'cv', 
               number = folds,
               summaryFunction=twoClassSummary,
               classProbs=TRUE)



```


## Parameters optimization


set initial values and try to optimize the model

```{r}
set.seed(1234)

#Run the model

rf_default <- train(sample~.,
                    data = train,
                    method = "rf",
                    metric = "ROC",
                    trControl = tc)

# Print the results

print(rf_default)
```



lets obtain the best mtry value 

```{r}

set.seed(1234)

tuneGrid <- expand.grid(.mtry = c(1: 15))

rf_mtry <- train(sample~.,
                 data = train,
                 method = "rf",
                 metric = "ROC",
                 tuneGrid = tuneGrid,
                 trControl = tc,
                 importance = TRUE,
                 nodesize = 14,
                 ntree = 50)
print(rf_mtry)

```


```{r}
#store the mtry to compare it with the other variable which we are going to 

best_mtry <- rf_mtry$bestTune$mtry 

print(best_mtry)
```


next lets train the maxnodes

```{r}
set.seed(1234)

store_maxnode <- list()
tuneGrid <- expand.grid(.mtry = best_mtry)
for (maxnodes in c(5:15)) {
  set.seed(1234)
  rf_maxnode <- train(sample~.,
                      data = train,
                      method = "rf",
                      metric = "ROC",
                      tuneGrid = tuneGrid,
                      trControl = tc,
                      importance = TRUE,
                      nodesize = 14,
                      maxnodes = maxnodes,
                      ntree = 50)
  current_iteration <- toString(maxnodes)
  store_maxnode[[current_iteration]] <- rf_maxnode
}
results_mtry <- resamples(store_maxnode)

summary(results_mtry)

```

lets try increasing the node and see if there is any change.

```{r}

set.seed(1234)

store_maxnode <- list()
tuneGrid <- expand.grid(.mtry = best_mtry)
for (maxnodes in c(20: 30)) {
  set.seed(1234)
  rf_maxnode <- train(sample~.,
                      data = train,
                      method = "rf",
                      metric = "ROC",
                      tuneGrid = tuneGrid,
                      trControl = tc,
                      importance = TRUE,
                      nodesize = 14,
                      maxnodes = maxnodes,
                      ntree = 50)
  key <- toString(maxnodes)
  store_maxnode[[key]] <- rf_maxnode
}
results_node <- resamples(store_maxnode)

summary(results_node)


```


max trees

```{r}
set.seed(1234)

store_maxtrees <- list()
for (ntree in c(10,15,20,25,30,35,40,45,50,55,100,150,200,250, 300,350,400)) {
  set.seed(5678)
  rf_maxtrees <- train(sample~.,
                       data = train,
                       method = "rf",
                       metric = "ROC",
                       tuneGrid = tuneGrid,
                       trControl = tc,
                       importance = TRUE,
                       nodesize = 14,
                       maxnodes = 5,
                       ntree = ntree)
  key <- toString(ntree)
  store_maxtrees[[key]] <- rf_maxtrees
}

results_tree <- resamples(store_maxtrees)

summary(results_tree)
```
## Optimized model

```{r}
set.seed(1234)

fit_rf <- train(sample~.,
                data= train,
                method = "rf",
                metric = "ROC",
                tuneGrid = tuneGrid,
                trControl = tc,
                importance = TRUE,
                nodesize = 14,
                ntree = 45,
                maxnodes = 6)

print(fit_rf$bestTune)
```

## Test the model

Test set

```{r}
set.seed(1234)

probs_test <- predict(fit_rf,test,type="prob")

test$sample = as.factor(test$sample)

gbm.ROC_test<- roc(predictor=probs_test$steatosis1,
                  response=test$sample)

table(test$sample)

print(gbm.ROC_test[["auc"]])

```

## Validate the model

Validation set 

```{r}
set.seed(1234)

probs_val <- predict(fit_rf,validation,type="prob")

validation$sample=as.factor(validation$sample)

gbm.ROC_val<- roc(predictor=probs_val$steatosis1,
               response=validation$sample)

table(validation$sample)

print(gbm.ROC_val[["auc"]])

```

## Plot the ROC-AUC curve

plot the curve 

```{r}
plot(gbm.ROC_test,col ="darkred",main="Random Forest ROC curve",
     col.lab="black", cex.lab=1.5)
text(x = 0.5274354,y = 0.7476923,label="AUC:0.74",cex=1)

plot(gbm.ROC_val,col ="darkgreen",add=T)
text(x = 0.7445764,y =0.8701577,label="AUC:0.76",cex=1)

legend(x = "bottomright", 
       c('Test-No.S0:190,No.S1:67','Validation-No.S0:120,No.S1:21','No.S0-Number of steatosis0',
         'No.S1-Number of steatosis1'),lty=c(1,1),
       lwd=c(2,2),col=c('darkred','darkgreen','white','white'))


```

session information

```{r}
sessionInfo()
```

